Survival in chessland

### 2020-03-09 trunk/chess/papers/survival.tex
Some work on longest paper/code. Shaved one slack move off, so feeling pretty confident in the argument that this is the longest...

  Since this is one of the shortest possible routes to a draw, I
  hereby dub this line the ``Fool's Draw,'' by analogy with the Fool's
  Mate. In this line all pieces survive, which is anyway humane and
  also advantageous in the case that you are someone is simultaneously
  being one of the chesspieces to the death!

@@

\begin{itemize}
\item Black and white are probably not significantly different. That
  is, the a1 and a8 rooks probably have about the same survival
  chances (it's known that white has a slight statistical
  advantage~\cite{chesstego} but it is probably only around 1\%). So
  these guesses will be written about white's pieces.
\item The d2 and e2 pawns are very active in common openings, and
  are frequently captured as part of those openings. I think they
  are the least likely to survive overall.
\item Bishops and knights are often involved in the opening and
  midgame, and often exchanged nonchalantly. I think they all have
  relatively low survival chances.
\item Although the queen is very valuable, a queen exchange is often
  forced for games that enter the endgame.
\item Rooks tend to be late-game pieces, because they are difficult
  to get out of their corners (and at most one can be activated
  by the fastest method, castling) and are relatively valuable.
\item This leaves the non-central pawns. These are the hardest to
  predict, and they are hard to think about (at least for me) because
  when e.g. the a2 pawn recaptures the b3 pawn that it supported, I
  just think of this as the b3 pawn. Of these pawns, b2 and g2 are
  somewhat weak because they are undefended once the bishop is
  developed (cf. the famous ``poison pawn'' at b2). On the other hand,
  in the fianchetto configuration, this pawn is very strong and often
  survives the entire game without leaving the third rank. Since pawn
  chains usually progress towards the middle of the board, the a2 pawn
  is more likely to be supporting than supported. This both leaves it
  weak to capture, but prone to recapturing. Outside pawns block one's
  own rook, although for this same reason they often clear the file by
  capturing (and so survive). They are also commonly used to push into
  a well-defended king's territory (e.g. in the fianchetto); kingside
  castling is more common, so this means that the h pawns are often
  lost to this fate.
\end{itemize}

@@

Like all good scientific research, I explicitly compare the actual
results to the hypotheses gathered before the experiment; this is a
hygenic and humbling exercise. Figure~\ref{fig:buddies} compares the
ranking across all games (slice {\bf Actual}) to the author author
(slice {\bf Tom}) and his drinking buddies (slices {\bf Ben}, {\bf
  Jim}, {\bf David}, {\bf William}). There are a number of different
reasonable ways to measure the accuracy of this type of position; a
very simple one is the sum of the absolute differences in rank for
each piece (e.g. if in one ranking \king is \#3, and in the other \#5,
then this contributes 2 to the total error). By this metric, Ben has
the best prediction (98 error), followed by David (138) and Tom (148).
Tom and David had the most similar predictions (116) and Jim and
William the most different (312). The expected error between two
completely random permutations is about 341, so all of these guesses
are significantly better than chance. Note in the actual ranking, many
pieces have very similar survival probabilities, and many guesses are
ambivalent about groups of pieces. Weighting each rank difference
equally is therefore an oversimplification. It would have been better
to ask each participant to give probabilities, as David did; this
would give us more sensitive error metrics and more opportunities to
spend the afternoon making visualizations.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{slices}
  \caption{
    Ranks and survival probabilities for different subsets of games.
    The same piece in adjacent rows is connected to highlight
    differences in the ranking, as before. The time formats all
    exhibit similar ranking with only small perturbations. Games
    including a titled player (rightmost column) are the most
    different, although we have far fewer samples in this set,
    so variance becomes significant.
  } \label{fig:slices}
\end{figure}

>>>>

  Since this is one of the shortest possible routes to a draw, I
  hereby dub this line the ``Fool's Draw,'' by analogy with the Fool's
  Mate. In this line all pieces survive, which is anyway humane and
  also advantageous in the case that you or someone is simultaneously
  being one of the chesspieces to the death!

@@

\begin{itemize}
\item Black and white are probably not substantially different. That
  is, the a1 and a8 rooks probably have about the same survival
  chances (it's known that white has a slight statistical
  advantage~\cite{chesstego} but it is probably only around 1\%). So
  these guesses will be written about white's pieces.
\item The d2 and e2 pawns are very active in common openings, and
  are frequently captured as part of those openings. I think they
  are the least likely to survive overall.
\item Bishops and knights are often involved in the opening and
  midgame, and often exchanged nonchalantly. I think they all have
  relatively low survival chances.
\item Although the queen is very valuable, a queen exchange is often
  forced for games that enter the endgame.
\item Rooks tend to be late-game pieces, because they are difficult
  to get out of their corners (and at most one can be activated
  by the fastest method, castling) and are relatively valuable.
\item This leaves the non-central pawns. These are the hardest to
  predict, and they are hard to think about (at least for me) because
  when e.g. the a2 pawn recaptures the b3 pawn that it supported, I
  just think of this as the b3 pawn. Of these pawns, b2 and g2 are
  somewhat weak because they are undefended once the bishop is
  developed (cf. the famous ``poison pawn'' at b2). On the other hand,
  in the fianchetto configuration, this pawn is very strong and often
  survives the entire game without leaving the third rank. Since pawn
  chains usually progress towards the middle of the board, the a2 pawn
  is more likely to be supporting than supported. This both leaves it
  weak to capture, but prone to recapturing. Outside pawns block one's
  own rook, although for this same reason they often clear the file by
  capturing (and so survive). They are also commonly used to push into
  a well-defended king's territory (e.g. in the fianchetto); kingside
  castling is more common, so this means that the h pawns are often
  lost to this fate.
\end{itemize}

@@

Like all good scientific research, I explicitly compare the actual
results to the hypotheses gathered before the experiment; this is an
hygenic and humbling exercise. Figure~\ref{fig:buddies} compares the
ranking across all games (slice {\bf Actual}) to the author author
(slice {\bf Tom}) and his drinking buddies (slices {\bf Ben}, {\bf
  Jim}, {\bf David}, {\bf William}). There are a number of different
reasonable ways to measure the accuracy of this type of position; a
very simple one is the sum of the absolute differences in rank for
each piece (e.g. if in one ranking \king is \#3, and in the other \#5,
then this contributes 2 to the total error). By this metric, Ben has
the best prediction (98 error), followed by David (138) and Tom (148).
Tom and David had the most similar predictions (116) and Jim and
William the most different (312). The expected error between two
completely random permutations is about 341, so all of these guesses
are significantly better than chance. Note in the actual ranking, many
pieces have very similar survival probabilities, and many guesses are
ambivalent about groups of pieces. Weighting each rank difference
equally is therefore an oversimplification. It would have been better
to ask each participant to give probabilities, as David did; this
would give us more sensitive error metrics and more opportunities to
spend the afternoon making visualizations.

@@

\begin{figure}[t]
  \includegraphics[width=\linewidth]{slices}
  \caption{
    Ranks and survival probabilities for different subsets of games.
    The same piece in adjacent rows is connected to highlight
    differences in the ranking, as before. The time formats all
    exhibit similar ranking with only small perturbations. Games
    including a titled player (rightmost column) are the most
    different, although we have far fewer samples in this set,
    so variance becomes significant.
  } \label{fig:slices}
\end{figure}
